---
title: "Intro to Spatial Data in R"
author: "Gaby Sanchez Ortiz"
date: "2023-01-25"
output: html_document
---

## Set up

```{r setup, include=FALSE}
source("setup.R")
```

## Vector data

Using tigris to work with polygons and lines

```{r polygons and lines}
# Downloading county boundary shapefile for CO
counties <- counties(state = "CO")

# Importing roads shapefile of Larimer county
roads <- roads(state = "CO", county = "Larimer")
```

Using tmap to create quick static and interactive maps

```{r mode and maps}
# View mode is make interactive plots
# Any plots created after the tmap_mode command, will be in the selected mode
tmap_mode("view")

# Making quick thematic maps (qtm) 
qtm(counties) +
  qtm(roads)

# Selecting the type of spatial layer using tm_shape()
tm_shape(counties) +
  tm_polygons() 
tm_shape(roads) +
  tm_lines()
```

Using sf (simple features) which represents spatial datas as a data.frame or tibble with a geometry column

```{r selecting Poudre Canyon Hwy}
# Filtering roads data for just Poudre Canyon Highway
roads %>%
  filter(FULLNAME == "Poudre Canyon Hwy") -> poudre_hwy
```

When working with spatial data, you start with some sort of excel or cvs file that is not in sf format, so you have to transform it.

```{r converting dataframes to sf}
# Creating a dataframe with points along the Poudre highway
poudre_points <- data.frame(name = c("Mishawaka", "Rustic", "Blue Lake Trailhead"),
                            long = c(-105.35634, -105.58159, -105.85563),
                            lat = c(40.68752, 40.69687, 40.57960))

# Convert the dataframe to sf format 
# a few notes: crs is coordinate reference system; also, x (long) goes first, y (lat) goes second
poudre_points_sf <- st_as_sf(poudre_points, coords = c("long", "lat"), crs = 4326)

qtm(poudre_hwy) +
  qtm(poudre_points_sf)
```

Working with spatial data requires a coordinate reference system. There are various ways to extract crs.

```{r crs}
# looking for crs in the metadata
counties # when we run we can see the geodetic crs is NAD83

# we can also use st_crs() to get more detailed information  
st_crs(counties)

# Checking to see if two objects have the same crs
st_crs(counties) == st_crs(poudre_points_sf) # they're not, so outcome is FALSE

# If they're not, you can re-project one based on the other's crs
poudre_points_prj <- st_transform(poudre_points_sf, st_crs(counties))

# The outcome should be TRUE now
st_crs(poudre_points_prj) == st_crs(counties)
```

## Raster data

Using elevatr to work with elevation data

```{r elevatr}
# get_elev_raster() needs spatial object to get the extent of elevation data (boundary basically) and the zoom level which is just the resolution
elevation <- get_elev_raster(counties, z = 7)

qtm(elevation) # gives default categorical symbology

# to change to continuous symbology
tm_shape(elevation) +
  tm_raster(style = "cont", title = "Elevation (m)")

# printing object gives detailed information about resolution, extent, crs, and file name
elevation
```

Working with raster data using terra

```{r terra}
# elevation raster has to be in proper format for terra (which is not done with elevatr) 

# convert RasterLayer object to terra SpatRaster object using rast()
elevation <- rast(elevation)

# changing the name
names(elevation) <- "Elevation"

# using crop() function to crop the raster to roads shapefile (Larimer county) and ext() function to get just the extent of the Poudre hwy
elevation_crop <- crop(elevation, ext(roads))

tm_shape(elevation_crop) +
  tm_raster(style = "cont")

# Creates a final map with all spatial data
tm_shape(elevation, bbox = st_bbox(poudre_hwy)) +
  tm_raster(style = "cont", title = "Elevation (m)") +
tm_shape(poudre_hwy) +
  tm_lines() +
tm_shape(poudre_points_prj) +
  tm_dots(size = 0.2)
```

## Reading and Writing Spatial Data

Writing spatial data

```{r writing}
# save vector data
write_sf(poudre_hwy, "data/poudre_hwy.shp")

write_sf(poudre_points_prj, "data/poudre_points.shp")

# save raster data
writeRaster(elevation_crop, "data/elevation_larimer.tif")

# saving variables as .RData files
save(poudre_hwy, poudre_points_prj, file = "data/spatial_object.RData")

# removing giles from environment
rm(poudre_hwy, poudre_points_prj)

# loading it back in
load("data/spatial_object.RData")
```

Reading spatial data

```{r reading}
# read shapefiles
read_sf("data/poudre_hwy.shp")

# read rasters
rast("data/elevation_larimer.tif")
```

## Exercises

1.  Filter out the counties data set to only include Larimer, Denver, and Pueblo counties.

```{r filter counties}
counties %>%
  filter(NAME == "Larimer" | NAME == "Denver" | NAME == "Pueblo") -> three_counties
```

2.  Make a map of the counties data colored by county area. Make a second map of counties colored by their total area of water.

```{r maps}
# map of counties data colored by county area
qtm(counties)

tm_shape(counties) +
  tm_polygons(col = "ALAND")

# map of counties data colored by area of water
tm_shape(counties) +
  tm_polygons(col = "AWATER")
```

3.  Make a bar plot comparing the elevation of your 3 points in the Poudre Canyon.

```{r barplot}

```

4.  Why are there 4 features in our Poudre Canyon Highway variable instead of 1.

Because the highway is cut into separate lines that are in slightly different places spatially.
