---
title: "Intro to R and R Markdown"
author: "Gaby Sanchez Ortiz"
date: "2023-01-20"
output: html_document
---

## Set up

Calls back to the script where we installed and read in libraries

```{r}
source("setup.R")
data("penguins")
```

## Exploring data types

Looking at class() and str() functions

```{r, datatypes}
class(penguins)
str(penguins)

class(penguins$species)
str(penguins$species)
```

## Creating Lists

```{r}
myList <- list("apple", 1993, FALSE, penguins)
str(myList)

# Nesting lists
list(myList, list("more stuff here", list("and more")))

#Using names() 
names(myList) <- c("fruit", "year", "logic", "data")
names(myList)
```

## Indexing

```{r}
# use double brackets for lists
myList[[1]]

myList[["data"]]

# for vectors use single brackets
myVector <- c("apple", "banana", "pear")
myVector[2]

# Index data frames
# dataframe[row(s), columns()]
penguins[1:5, 2]

penguins[1:5, "island"]

penguins[1, 1:5]

penguins[1:5, c("species", "sex")]

penguins[penguins$sex=='female',]

# $ for a single column
penguins$species

```

## Indexing Exercises

1.  Why don't the following lines of codes work? Tweak each one so the code runs.

```{r}
# original was myList["Fruit]
# to correct, you need to add double brackets and keep the name as you originally wrote it
myList[["fruit"]]

# original was penguins$flipper_lenght_mm
# to correct you need to make sure you are spelling the column names correctly
penguins$flipper_length_mm

# original was penguins[island=='Dream',]
# to correct you need to specify that the column you are looking at is in the penguins dataframe
penguins[penguins$island=='Dream',]
```

2.  How many species are in the penguins dataset? What islands were collected for?

```{r}
# Number of species in penguins dataset
numSpecies <- unique(penguins$species)
length(numSpecies)

# Individual islands collected for
unique(penguins$island)
```

3.  Use indexing to create a new dataframe that has only 3 columns: species, island, flipper length, and subset all rows for just the 'Dream' island.

```{r}
# Indexing for just three columns and just the Dream island
penguins[penguins$island=='Dream', c("species", "island", "flipper_length_mm")]
```

4.  Use indexing and the mean() function to find the average flipper length for the Adelie species on Dream island.

```{r}
# Filters penguins df to only include all of the rows of Adelie penguins in Dream island
AdelieDream <-penguins[penguins$species=="Adelie" & penguins$island=="Dream",]

# Takes the average of their flipper length but does not take into account the rows with NA
mean(AdelieDream$flipper_length_mm, na.rm=TRUE)
```

## Exploring the dplyr package

Subset rows with filter()

```{r}
filter(penguins, species == "Adelie")

filter(penguins, species != "Adelie")

filter(penguins, island %in% c("Dream", "Torgersen") & !is.na(bill_length_mm))
```

Select columns with select()

```{r}
# Select two specific variables
select(penguins, species, sex)

# Select a range of variables
select(penguins, species:flipper_length_mm)

# Rename columns within select
select(penguins, genus = species, island)

# Select column variables that are recorded in mm
select(penguins, contains("mm"))
```

Create new variables with mutate()

```{r}
# New variable that calculates bill length in cm
mutate(penguins, bill_length_cm = bill_length_mm/10)

# mutate based on conditional statements
mutate(penguins, species_sex = if_else(sex == 'male', paste0(species,"_m"), paste0(species, "_f")))
```

## Pipe %\>% Exercises

1.  Reorder the variables in penguins so that year is the first column followed by the rest

```{r}
penguins %>%
  select(year, everything())
```

2.  Create a new column called 'size_group' where individuals with body mass greater than the overall average are called 'large' and those smaller are called 'small'

```{r}
penguins %>%
  mutate(size_group = if_else(body_mass_g > mean(body_mass_g, na.rm = T), paste0('large'), paste0('small')))
```

3.  Find out which year for each species were individuals on average the largest according to body mass.

```{r}
penguins %>%
  group_by(species,year) %>%
  summarise(year_largest_avg = mean(body_mass_g, na.rm = T))
# would i need to group and resummarise?
```

4.  You want to filter data for years that are NOT in a vector of given years, but this code doesn't work. Tweak it so that it does.

```{r}
# original was:
# penguins %>%
#   filter(year !%in% c(2008, 2009))

penguins %>%
  filter(!year %in% c(2008, 2009))
```

## Visualize

Exploring histograms

```{r}
ggplot(penguins) +
  geom_histogram(mapping = aes(x = flipper_length_mm))

# flipper length by species
ggplot(penguins) +
  geom_histogram(aes(x = flipper_length_mm, fill = species), alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4"))

# same example as previous, but in different plots
ggplot(penguins) +
  geom_histogram(aes(x = flipper_length_mm, fill = species), alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  facet_wrap(~species)
```

Bar plots

```{r}
ggplot(penguins) +
  geom_bar(mapping = aes(x = island, fill = species))

# Creating different bar plots for every species and plotting by island

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"), 
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Scatter plots

```{r}
ggplot(penguins) +
  geom_point(mapping = aes(x = body_mass_g, y = flipper_length_mm, color = species))
```

## Visualization Exercises

1.  Make a bar plot showing the average flipper length for each species.

```{r}
ggplot(penguins) +
  geom_bar(mapping = aes(x = mean(flipper_length_mm, na.rm = T), fill = species))
```

2.  Make a scatter plot of bill length compared to bill depth but only for observations on Dream island.

```{r}
penguins %>%
  filter(island == 'Dream') %>%
  ggplot() +
    geom_point(mapping = aes(x = bill_length_mm, y = bill_depth_mm))
```
